---
title: "Flight Delay Prediction Analysis"
format: docx
editor: 
  markdown: 
    wrap: 72
---

# Introduction 

## Background

Flight delays are undeniably one of the most frustrating aspects of
travel, particularly for college students balancing tight schedules.
Juggling academic responsibilities, social commitment, and possibly part
time jobs or internships already leaves us with precious little time to
space. Despite airline’s marketing and promises of efficiency and
punctuality, timely arrivals and departures often seem to elude us
during peak travel seasons. Exploring the various factors influencing
flight delays, including weather conditions, airport congestion, and
airline operations, could provide great insight into this persistent
challenge. As a result, we decided to investigate this issue further,
utilizing the January Flight Delay Prediction dataset from the Bureau of
Transportation Statistics (BTS). This comprehensive dataset specifically
focuses on January 2019 Ontime flight information, offering us an
opportunity to analyze patterns and potentially develop strategies to
better anticipate delays. For college students with limited flexibility
in our schedules, understanding these patterns could significantly
improve our travel experiences and help us make more informed decisions
about when and how to plan our journeys.

## Our Dataset

As mentioned before, the January Flight Delay Prediction dataset
originates from the Bureau of Transportation Statistics (BTS),
specifically focusing on the January 2019 Ontime dataset. This
comprehensive dataset contains 583,985 rows and 21 columns (583,985
observations and 21 variables) related to flight information and delay
patterns. The variables are as follows:

-   DAY_OF_MONTH - day of month on which the flight occurred
-   DAY_OF_WEEK - day of week on which the flight occurred
-   OP_UNIQUE CARRIER - a unique code assigned to the operating airline
-   OP_CARRIER_AIRLINE_ID - a unique numerical ID for the operating
    airline
-   OP_CARRIER - the operating carrier code
-   TAIL_NUM - the tail number of the aircraft, unique for each plane
-   OP_CARRIER_FL_NUM - the flight number assigned by the operating
    carrier
-   ORIGIN_AIRPORT_ID - a unique numerical ID for the origin airport
-   ORIGIN_AIRPORT_SEQ_ID - a sequential ID for the origin airport
-   ORIGIN - the IATA code for the origin airport
-   DEST_AIRPORT_ID - a unique numerical ID for the destination airport
-   DEST_AIRPORT_SEQ_ID - a sequential ID for the destination airport
-   DEST - the IATA code for the destination airport
-   DEP_TIME - the scheduled departure time for the flight
-   DEP_DEL15 - a binary indicator showing whether the flight departure
    was delayed by 15 minutes or more (0 = not delayed, 1 = delayed)
-   DEP_TIME_BLK - the scheduled departure time block, broad categories
-   ARR_TIME - the scheduled arrival time of the flight
-   ARR_DEL15 - a binary indicator showing whether the flight arrival
    was delayed by 15 minutes or more (0 = not delayed, 1 = delayed)
-   CANCELLED - a binary indicator showing whether the flight was
    cancelled (0 = not cancelled, 1 = cancelled)
-   DIVERTED - a binary indicator showing whether the flight was
    diverted to an alternate airport (1 = yes, 0 = no)
-   DISTANCE - the distance between the origin and destination airport

## Our Question

Our primary research question is: **What factors contribute to airline
departure delays, and can we accurately predict a delay of 15 minutes or
more using information available prior to departure?**

# Data Cleaning 

To prepare for our predictive modeling of flight delays, we needed to clean and format our dataset. After carefully evaluating the original dataset containing 583,985 observations and 21 variables, we identified which features would be most useful for our analysis and which might interfere with model accuracy.
With this goal in mind, we selected key variables that would be known prior to departure and dropped those that were redundant, irrelevant, or would create data leakage in our predictive models. The summary statistics of our original dataset revealed the range and distribution of variables such as departure times, flight distances, and our target variable DEP_DEL15 (indicating whether a flight was delayed by 15 minutes or more).

We focused our cleaned dataset on the following variables:

- DAY_OF_WEEK: Day of week on which the flight occurred
- DAY_OF_MONTH: Day of month on which the flight occurred
- OP_CARRIER: The operating carrier code
- ORIGIN: The IATA code for the origin airport
- DEST: The IATA code for the destination airport
- DEP_TIME: The scheduled departure time for the flight
- DEP_TIME_BLK: The scheduled departure time block
- DISTANCE: The distance between origin and destination airports
- DEP_DEL15: Our target variable indicating whether departure was delayed (1) or not (0)

We also removed rows with missing values using na.omit() to ensure data quality. To improve model performance and prevent overfitting, we reduced the factor levels for certain categorical variables (ORIGIN, DEST, and OP_CARRIER) by keeping only the top 10 most frequent values and grouping the rest as "OTHER."


```{r}
# Load the dataset
flight_data <- read.csv("Jan_2019_ontime.csv")

# Data cleaning - select relevant features from our cleaned dataset
flight_clean <- flight_data[c("DAY_OF_WEEK", "DAY_OF_MONTH", "OP_CARRIER", "ORIGIN", 
                             "DEST", "DEP_TIME", "DEP_TIME_BLK", "DISTANCE", "DEP_DEL15")]

# Remove rows with NA
flight_clean <- na.omit(flight_clean)

# Reduce factor levels to top 10 for better performance
top_origin <- names(sort(table(flight_clean$ORIGIN), decreasing = TRUE)[1:10])
flight_clean$ORIGIN <- ifelse(flight_clean$ORIGIN %in% top_origin, flight_clean$ORIGIN, "OTHER")

top_dest <- names(sort(table(flight_clean$DEST), decreasing = TRUE)[1:10])
flight_clean$DEST <- ifelse(flight_clean$DEST %in% top_dest, flight_clean$DEST, "OTHER")

top_carrier <- names(sort(table(flight_clean$OP_CARRIER), decreasing = TRUE)[1:10])
flight_clean$OP_CARRIER <- ifelse(flight_clean$OP_CARRIER %in% top_carrier, flight_clean$OP_CARRIER, "OTHER")

# Summary statistics
summary(flight_data)
```

From the summary statistics, we can observe that approximately 17.4% of flights experienced delays (mean of DEP_DEL15 = 0.174), making this a somewhat imbalanced classification problem that we'll need to address in our modeling approach.

Additionally, we properly converted all categorical variables to factors to ensure appropriate handling in our models:

- DAY_OF_WEEK
- DAY_OF_MONTH
- OP_CARRIER
- ORIGIN
- DEST
- DEP_TIME_BLK
- DEP_DEL15 (our target variable)

```{r}
flight_clean$DAY_OF_WEEK <- as.factor(flight_clean$DAY_OF_WEEK)
flight_clean$DAY_OF_MONTH <- as.factor(flight_clean$DAY_OF_MONTH)
flight_clean$OP_CARRIER <- as.factor(flight_clean$OP_CARRIER)
flight_clean$ORIGIN <- as.factor(flight_clean$ORIGIN)
flight_clean$DEST <- as.factor(flight_clean$DEST)
flight_clean$DEP_TIME_BLK <- as.factor(flight_clean$DEP_TIME_BLK)
flight_clean$DEP_DEL15 <- as.factor(flight_clean$DEP_DEL15)
```

## Our Chosen Approaches

With our data cleaned and our response variable (DEP_DEL15) defined as
binary (0 = no delay, 1 = delay), we shifted our focus toward
identifying the most appropriate models for classification. Since we are
not predicting a continuous outcome, regression models like linear
regression were not suitable. Instead, we selected four classification
models that would allow us to predict the likelihood of a flight delay
and analyze the most influential predictors: Logistic Regression,
Decision Tree, Random Forest, and Linear Discriminant Analysis (LDA).

First thing's first, divide data into training and test:

```{r}
# Split into training (75%) and test set (25%)
library(caret)
set.seed(42)
trainIndex <- createDataPartition(flight_clean$DEP_DEL15, p = 0.75, list = FALSE)
train_data <- flight_clean[trainIndex, ]
test_data <- flight_clean[-trainIndex, ]

# Address class imbalance using ROSE instead of SMOTE
library(ROSE)
balanced_train <- ROSE(DEP_DEL15 ~ ., data = train_data, seed = 123)$data
```

# Our Methods

## Logistic Regression Model

## Why We Chose this Model

We began with logistic regression as our baseline classification model
because it is widely used for binary classification problems and
provides clear interpretations of predictor coefficients. Logistic
regression models the probability of a flight being delayed based on a
logistic function and is especially useful when the relationship between
predictors and the log-odds of the response is approximately linear.

## Our Formula

In the given equation, $p$, represents the probability of a flight being
delayed, $B_0$ represents the model intercept, and $B_i$ for $i=1,...,n$
represents the coefficients of the predictor variables, where:

-   $B_1$ represents the coefficient for DAY_OF_WEEK
-   $B_2$ represents the coefficient for OP_CARRIER
-   $B_3$ represents the coefficient for ORIGIN_AIRPORT_ID
-   $B_4$ represents the coefficient for DEST_AIRPORT_ID
-   $B_5$ represents the coefficient for DEP_TIME_BLK
-   $B_6$ represents the coefficient for DISTANCE

So, the logistic regression model equation is expressed as:

$y = \frac{e^{B_0 + B_1(\text{DAY_OF_WEEK}) + B_2(\text{OP_CARRIER}) + B_3(\text{ORIGIN_AIRPORT_ID}) + B_4(\text{DEST_AIRPORT_ID}) + B_5(\text{DEP_TIME_BLK}) + B_6(\text{DISTANCE})}}{1 + e^{B_0 + B_1(\text{DAY_OF_WEEK}) + B_2(\text{OP_CARRIER}) + B_3(\text{ORIGIN_AIRPORT_ID}) + B_4(\text{DEST_AIRPORT_ID}) + B_5(\text{DEP_TIME_BLK}) + B_6(\text{DISTANCE})}}$


```{r}
logistic_model <- glm(DEP_DEL15 ~ ., data = balanced_train, family = binomial)
summary(logistic_model)
```

## Model Interpretation

### Intercept

The intercept ($B_0$) is -3.032, which means that when all predictor variables are at their reference levels (i.e., when the day of the week, airline, airport, departure time block, and distance are at their baseline), the log-odds of a flight being delayed is -3.032. This suggests a relatively low probability of delay under these conditions, but this is just the baseline.

### Significant Predictors

- DAY_OF_WEEK: Days like Thursday (DAY_OF_WEEK4) increase the likelihood of delays, while Saturday (DAY_OF_WEEK6) reduces it
- OP_CARRIER: American Airlines (OP_CARRIERAA) has a lower chance of delays, while JetBlue (OP_CARRIERB6) has a higher chance
- ORIGIN_AIRPORT_ID & DEST_AIRPORT_ID: Both airports have small positive coefficients, indicating a slight increase in delays based on the airport
- DEP_TIME_BLK: Later departure times, like 10:00-11:00 AM (DEP_TIME_BLK1000-1059), significantly increase delays
- DISTANCE: Longer flights have a small increased likelihood of delays

## Further Analysis - Confusion Matrix

```{r}
# Predict on test data
logistic_pred <- predict(logistic_model, newdata = test_data, type = "response")
logistic_pred_class <- ifelse(logistic_pred > 0.5, 1, 0)
logistic_pred_class <- as.factor(logistic_pred_class)

# Confusion Matrix 
conf_matrix <- confusionMatrix(logistic_pred_class, test_data$DEP_DEL15)
print(conf_matrix)
```
The confusion matrix shows the following results:

- True Negatives (TN): 79,714 flights were correctly predicted as not delayed.
- False Positives (FP): 37,461 flights were incorrectly predicted as delayed when they were not.
- False Negatives (FN): 7,314 flights were incorrectly predicted as not delayed when they actually were delayed.
- True Positives (TP): 17,417 flights were correctly predicted as delayed.

From this, we can observe that our logistic regression model achieved an accuracy of 68.45%, which is lower than the No Information Rate of 82.57%. This suggests that simply predicting the majority class (not delayed) would actually yield higher accuracy than our model. The Kappa value of 0.2597 indicates fair agreement beyond chance, but there's significant room for improvement.

The model demonstrates balanced sensitivity (68.03%) and specificity (70.43%), indicating it performs similarly at identifying both delayed and non-delayed flights. However, the positive predictive value of 91.60% for non-delayed flights versus the negative predictive value of 31.74% for delayed flights shows that the model is much better at correctly identifying non-delayed flights when it predicts them. This imbalance suggests that despite our efforts to address class imbalance with ROSE, the model still struggles with accurately predicting the minority class (delayed flights).

## Tree-Based Models

### Why We Chose this Model

We chose a decision tree classifier as an additional modeling approach due to its interpretability and ability to capture complex, non-linear relationships between flight attributes and delays. Unlike logistic regression, which assumes linear relationships between features and the target, decision trees can reveal key decision splits that lead to delays. However, decision trees are prone to overfitting, so we applied pruning techniques and also experimented with ensemble methods like Random Forest to improve generalization.

### Our Formula
DEP_DEL15 ~ DAY_OF_MONTH + DAY_OF_WEEK + OP_CARRIER + ORIGIN + DEST + DEP_TIME_BLK + DISTANCE

### Decision Tree Analysis

Using the rpart package in R, we constructed a classification tree to predict flight delays. The decision tree creates binary splits in the data based on feature values, recursively partitioning the data to maximize homogeneity in each resulting node.

```{r}
# Continuing with the same data preparation steps from the logistic regression section
library(rpart)
library(rpart.plot)
library(caret)

# Building the decision tree model
tree_model <- rpart(DEP_DEL15 ~ DAY_OF_MONTH + DAY_OF_WEEK + OP_CARRIER + ORIGIN + 
                   DEST + DEP_TIME_BLK + DISTANCE, 
                   data = train_data, method = "class",
                   control = rpart.control(minsplit = 20, cp = 0.001))

# Printing tree model
print(tree_model)
```

```{r}
# Visualize the decision tree
rpart.plot(tree_model, extra = 104, under = TRUE, box.palette = "RdBu", 
           fallen.leaves = TRUE, main = "Decision Tree for Flight Delay Prediction")
```

### Tree structure

The decision tree visualization shows the hierarchical structure of our flight delay prediction model. The tree branches out from the root node at the top, making binary splits based on the most informative features in our dataset. Each internal node represents a decision point based on a specific variable (such as departure time, airline carrier, or airport), while the leaf nodes at the bottom indicate the final prediction outcomes (0 for no delay, 1 for delay).


```{r}
# Making predictions on the test set
tree_pred <- predict(tree_model, newdata = test_data, type = "class")

# Confusion matrix
tree_conf_matrix <- confusionMatrix(tree_pred, test_data$DEP_DEL15)
print(tree_conf_matrix)

# Calculating accuracy, sensitivity, specificity
tree_accuracy <- tree_conf_matrix$overall["Accuracy"]
tree_sensitivity <- tree_conf_matrix$byClass["Sensitivity"]
tree_specificity <- tree_conf_matrix$byClass["Specificity"]

# Performance metrics
cat("Decision Tree Performance:\n")
cat("Accuracy:", tree_accuracy, "\n")
cat("Sensitivity (True Positive Rate):", tree_sensitivity, "\n")
cat("Specificity (True Negative Rate):", tree_specificity, "\n")
```
## Further Analysis - Confusion Matrix

The confusion matrix for the decision tree model shows:

True Negatives (TN): 115,896 flights correctly predicted as not delayed
False Positives (FP): 1,279 flights incorrectly predicted as delayed when they weren't
False Negatives (FN): 22,601 flights incorrectly predicted as not delayed when they were delayed
True Positives (TP): 2,130 flights correctly predicted as delayed

The model achieved 83.17% accuracy, slightly better than the No Information Rate of 82.57%. However, the low Kappa value (0.114) indicates poor agreement beyond what would be expected by chance.
While sensitivity is very high (98.91%), meaning the model excels at identifying non-delayed flights, the specificity is extremely low (8.61%), showing it struggles significantly with identifying truly delayed flights. This severe imbalance suggests the model is biased toward predicting the majority class (not delayed) and performs poorly on the minority class, despite achieving decent overall accuracy.

### Pruning to Avoid Overfitting
To prevent our decision tree from capturing noise in the training data, we implemented cost-complexity pruning. By examining the cross-validated error rates across different complexity parameters, we identified an optimal tree size that balances model complexity and predictive accuracy.

The pruned tree retained the most informative splits while eliminating branches that contributed little to prediction accuracy. This resulted in a more generalizable model that performs better on unseen data.

```{r}
# Examine pruning to avoid overfitting
printcp(tree_model)
plotcp(tree_model)

# Select optimal complexity parameter
optimal_cp <- tree_model$cptable[which.min(tree_model$cptable[,"xerror"]),"CP"]
cat("Optimal complexity parameter:", optimal_cp, "\n")

# Prune the tree
pruned_tree <- prune(tree_model, cp = optimal_cp)
```

```{r}
# Visualize the pruned tree
rpart.plot(pruned_tree, extra = 104, under = TRUE, box.palette = "RdBu", 
           fallen.leaves = TRUE, main = "Pruned Decision Tree for Flight Delay Prediction")
```

```{r}
# Make predictions with pruned tree
pruned_pred <- predict(pruned_tree, newdata = test_data, type = "class")

# Create confusion matrix for pruned tree
pruned_conf_matrix <- confusionMatrix(pruned_pred, test_data$DEP_DEL15)
print(pruned_conf_matrix)

# Compare performance before and after pruning
cat("Comparison of accuracy before and after pruning:\n")
cat("Original tree accuracy:", tree_accuracy, "\n")
cat("Pruned tree accuracy:", pruned_conf_matrix$overall["Accuracy"], "\n")
```


#### Performance Metrics
Our decision tree achieved 83.17% accuracy on the test set, with a sensitivity of 98.91% and specificity of 8.61%. Compared to logistic regression, the decision tree performed worse at identifying actual delayed flights, despite having higher overall accuracy. The extremely low specificity (8.61%) indicates that while the model is excellent at identifying non-delayed flights (98.91% sensitivity), it struggles significantly to correctly classify delayed flights. This severe imbalance suggests that the model is heavily biased toward predicting the majority class (non-delayed flights).

The Kappa value of 0.114 shows poor agreement beyond chance, indicating that despite the seemingly decent accuracy, most of this performance comes from simply predicting the majority class. The large gap between sensitivity and specificity demonstrates that the non-linear relationships captured by the tree were not particularly valuable for this prediction task when it comes to identifying the minority class of delayed flights.
